\ZpL: a p-adic precision package

### 2018-01-17 ZpL/ZpL.tex
First version of 3.1

\subsection{The precision Lemma}

@@

\begin{tabular}{|l|l|}
\verb?R = Qp(2,250)? & \verb?R = QpL(2,250)? \\
\verb?S = Zp(2,250)? & \verb?S = ZpL(2,250)? \\
\multicolumn{2}{|l|}{\tt A.<x> = PowerSeriesRing(R,32) } \\
\multicolumn{2}{|l|}{\tt B.<x> = PowerSeriesRing(S,32)) } \\
\multicolumn{2}{|l|}{\tt h,y =1+x+x\^{}3 , x+x\^{}2*A(B.random\_element()) } \\
\multicolumn{2}{|l|}{\tt g=y.derivative()/h(y) } \\
\multicolumn{2}{|l|}{\tt s=Newton\_Iteration\_Solver(g,h,5)} \\
\multicolumn{2}{|l|}{\tt print 250-(s[31].lift()-y[31].lift()).valuation(2)} \\
\hfill\verb?9? & \hfill\verb?4? \\
\end{tabular}
The theorical loss in precision for the coefficient of $t^{31}$
is $\lfloor log_2 (31)\rfloor=4$ and is then reached, whereas the
loss in precision for a naive implementation is clearly more severe.

@@

We can solve numerically 
the corresponding differential equation
using the algorithm of \cite{Lercier-Sirvent:08}
over $\mathbb{Q}_3$ with lattice precision.
Any generic implementation of the Newton
method of \cite{Lercier-Sirvent:08} is enough.

\begin{tabular}{|l|}
\verb?    R = QpL(3)? \\
\verb?    A.<x> = PolynomialRing(R)?\\
\verb?    m=67?\\
\verb?    g=1+(1/4)*x^2+x^6?\\
\verb?    h=1+((m^2)/4)*x^2+m^6* x^6?\\
\verb?    y=Lercier_Sirvent(g,h,64)?\\
\verb?    y[31]?\\
\verb?(467995903278061284942799/3^14 + O(3^36))*x^31?\\
\end{tabular}

>>>>

\subsection{The precision Lemma}
In \cite{caruso-roe-vaccon:14a}, we have suggested the 
use of lattices to represent the precision of elements in 
$K$-vector spaces.  It contrasts with the
\emph{coordinate-wise method} (of \textit{e.g.}
 \verb?Zp(5)?) that is of use in Sage prior to
 our package, and  
where the precision of an element is specified by giving the precision
of each coordinate separately and is updated after each basic
operation.

Consider a finite 
dimensional normed vector space $E$ 
defined over $K$. 
We use the notation $\Vert \cdot \Vert_E$ for the norm 
on $E$ and $B^-_E(r)$ (resp. $B^{\phantom -}_E(r)$) for the open (resp. 
closed) ball of radius $r$ centered at the origin. A \emph{lattice} $L \subset 
E$ is a sub-$\O_K$-module which generates $E$ over $K$. 
Because of ultrametricity, the balls $B^{\phantom 
-}_E(r)$ and $B^-_E(r)$ are examples of lattices. 
Lattices can be thought of as special neighborhoods of $0,$
 and therefore are good 
candidates to model precision data. Moreover, as revealed in 
\cite{caruso-roe-vaccon:14a}, they behave quite well under (strictly) 
differentiable maps:

\begin{prop}
\label{prop:precision}
Let $E$ and $F$ be two finite dimensional normed vector spaces over $K$ 
and $f : U \rightarrow F$ be a function defined on an open subset $U$ of 
$E$. We assume that $f$ is differentiable at some point $v_0 \in U$ and 
that the differential $f'(v_0)$ is surjective.
Then, for all $\rho \in (0, 1]$, there exists a positive real
number $\delta$ such that, for all $r \in (0, \delta)$, any lattice
$H$ such that $B^-_E(\rho r) \subset H \subset B^{\phantom -}_E(r)$ 
satisfies:
\begin{equation}
\label{eq:firstorder}
f(v_0 + H) = f(v_0) + f'(v_0) (H).
\end{equation}
\end{prop}

This proposition enables the \emph{lattice method} of tracking precision,
where the precision of the input is specified as a lattice $H$ and precision
is tracked via differentials of the steps within a given algorithm.
The equality sign in Eq.~\eqref{eq:firstorder} shows that this method
yields the optimum possible precision. 
We refer to \cite[\S 4.1]{caruso-roe-vaccon:14a} for a more complete 
exposition.
We emphasize that polynomials, rational fractions (outside of their poles)
and converging power series are strictly differentiable.

The lemma is more precise in the context of so-called
integral polynomial functions.
A function $f : E \to F$ is said to be
\emph{integral polynomial} if it is given by multivariate polynomial 
functions with coefficients in $\O_K$ in any (equivalently all) system 
of coordinates associated to a $\OK$-basis of $B_E(1)$.

\begin{prop}
\label{prop:precision2}
We keep the notation of Proposition \ref{prop:precision} and assume 
in addition that $f$ is integral polynomial. Let $C$ be a positive real
number such that $B_F(1) \subset f'(v_0)(B_E(C))$. 
Then Proposition \ref{prop:precision} holds with $\delta = C \cdot
\rho^{-1}$.
\end{prop}

We will use the following definition to measure the gap
between lattice and coordinate-wise methods. 
Suppose $E$ is equipped with a basis $(e_1, \ldots, e_n)$ and write $\pi_i : E \to Ke_i$ for the projections.

\begin{deftn} \label{def:diffused}
Let $H \subset E$ be a lattice. The number of 
\emph{diffused digits of precision} of $H$ is the length of
$H_0/H$ where $H_0 = \pi_1(H) \oplus \cdots \oplus \pi_n(H)$.
\end{deftn}
If $H$ represents the actual precision on some object, then
$H_0$ is the smallest diagonal lattice containing $H$.
Coordinate-wise methods, with one precision datum per coordinate,
corresponds to diagonal lattices.
Hence, they cannot yield a precision better than $H_0$, and
then $k$ provides a lower bound on the number of $p$-adic digits
gained by lattice methods over standard methods.

@@

\begin{tabular}{|l|l|}
\verb?R = Qp(2,250)? & \verb?R = QpL(2,250)? \\
\verb?S = Zp(2,250)? & \verb?S = ZpL(2,250)? \\
\multicolumn{2}{|l|}{\tt A.<x> = PowerSeriesRing(R,32) } \\
\multicolumn{2}{|l|}{\tt B.<x> = PowerSeriesRing(S,32)) } \\
\multicolumn{2}{|l|}{\tt h,y =1+x+x\^{}3 , x+x\^{}2*A(B.random\_element()) } \\
\multicolumn{2}{|l|}{\tt g=y.derivative()/h(y) } \\
\multicolumn{2}{|l|}{\tt s=Newton\_Iteration\_Solver(g,h,5)} \\
\multicolumn{2}{|l|}{\tt print 250-(s[31].lift()-y[31].lift()).valuation(2)} \\
\hfill\verb?9? & \hfill\verb?4? \\
\end{tabular}
The last line provides the loss in precision between the
original solution of the differential equation and its approximation
obtained through computation.
The theorical loss in precision for the coefficient of $t^{31}$
is $\lfloor log_2 (31)\rfloor=4$ and is then reached, whereas the
loss in precision for a naive implementation is clearly more severe.

@@

We can solve numerically 
the corresponding differential equation
using the algorithm of \cite{Lercier-Sirvent:08}
over $\mathbb{Q}_3$ with lattice precision.
We use an initial precision of 50 digits.
Any generic implementation of the Newton
method of \cite{Lercier-Sirvent:08} is enough.

\begin{tabular}{|l|}
\verb?    R = QpL(3,50)? \\
\verb?    A.<x> = PolynomialRing(R)?\\
\verb?    m=67?\\
\verb?    g=1+(1/4)*x^2+x^6?\\
\verb?    h=1+((m^2)/4)*x^2+m^6* x^6?\\
\verb?    y=Lercier_Sirvent(g,h,64)?\\
\verb?    y[31]?\\
\verb?(467995903278061284942799/3^14 + O(3^36))*x^31?\\
\end{tabular}
